---
title: "RepData_PeerAssessment1"
author: "Navin Bhatera"
date: "Monday, March 16, 2015"
output: html_document
---

This document contains the Peer Assignment 1 for Reproducible Research course as part of Data Science specialization at Coursera

Loading and Preprocessing data : 
```{r first,echo = TRUE}
library(ggplot2)
activity <- read.csv("activity.csv",header=T,colClasses=c("numeric", "character", "numeric"))
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")
activity$interval <- as.factor(activity$interval)
str(activity)
```

Histogram for the processed data (excluding NA's)

```{r second,echo=TRUE}
daysteps <- tapply(activity$steps, activity$date, FUN=sum, na.rm=T)
hist(daysteps,main="Number of steps taken per day", xlab="Steps", col="red",breaks=10)

```

The mean of steps taken per day is 

```{r third,echo=TRUE}
mean(daysteps, na.rm=T)
```

The median of steps taken per day is 

```{r fourth,echo=TRUE}
median(daysteps, na.rm=T)
```

Avergage daily actiity pattern is plotted below
```{r fifth, echo=TRUE}
daysteps_mean <- aggregate(x=list(steps=activity$steps), by = list(interval=activity$interval),
                           FUN=mean, na.rm=T)
daysteps_mean$interval <- as.integer(levels(daysteps_mean$interval)[daysteps_mean$interval])
ggplot(daysteps_mean, aes(x=interval, y=steps)) +   
    geom_line(color="red", size=1) +  
    labs(title="Average Daily Activity Pattern", x="Interval", y="Steps")
```

Max number of steps for a particular 5-min interval can be located at 

```{r sixth, echo=TRUE}
daysteps_mean[which.max(daysteps_mean$steps),]
```


New dataset creation: (using mean of the day as filler)

```{r seventh, echo=TRUE}
fillstep <- function(steps, int) {
  temp = 0
  if (!is.na(steps)) temp = steps
  else temp = daysteps_mean[daysteps_mean$interval == int, "steps"]
  return(temp)
  }
activity_fill <- activity
activity_fill$steps <- mapply(fillstep,activity_fill$steps,activity_fill$interval)
```

Histogram Plot with the new data set: (NA's replaced by Day Mean)

```{r eight, echo=TRUE}
daysteps_fill <- tapply(activity_fill$steps, activity_fill$date, FUN=sum, na.rm=T)
hist(daysteps_fill, main="Number of steps taken per day", xlab="Steps",col="blue",breaks=10)

```


The mean of steps taken per day is 

```{r ninth,echo=TRUE}
mean(daysteps_fill)
```

The median of steps taken per day is 

```{r tenth,echo=TRUE}
median(daysteps_fill)
```


Now evaluating the difference between weekdays and weekend activity pattern: 

```{r eleventh, echo=TRUE}
weekdaytype <- function(date){
  day <- weekdays(date)
  if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
    return("weekday") 
  else if (day %in% c("Saturday", "Sunday")) 
    return("weekend")
  else stop("invalid date")
  }
activity_fill$day <- sapply(activity_fill$date, FUN = weekdaytype)

daytype_mean <- aggregate(steps ~ interval + day, data = activity_fill, mean)
ggplot(daytype_mean, aes(as.numeric(interval), steps)) + geom_line(col="blue") + facet_grid(day ~ .) + labs(title="Average Daily Activity Pattern", x="Interval", y="Steps") + theme_bw()

```
